**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-06-12 submission/manuscript.Rmd
refs done

---
title: '**Machine learning identifies unique taxa differentiating proximal and distal
  human colonic microbiota**'
fontsize: 11pt
geometry: margin=1.0in
output:
  html_document: default
  pdf_document:
    includes: null
    keep_tex: yes
csl: plos.csl
bibliography: references.bib 
---
\vspace{35mm}

@@

Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies. A rare report of a matched-colonoscopy study that sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (cite). This group found that mucosa and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (Shobar 2016). Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites. Further, the differences were not great enough to overcome interpersonal differences between subjects. So, bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel. Thus our findings of specific bacteria at each site of the colon are strengthened by the lack of bowel preparation.

Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer. Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors. Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit (Baxter et al 2016) and it performed much better than FIT alone. Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (cite Niel fit paper). Our work already showed that the stool most accurately matched the samples from the left lumen. So we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities. But further studies would need to be done. Hard to get a complete set of matched samples from CRC patients though due to discomfort and insurance issues. One caveat of the previous study is that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective.

offer another model/schematic? what does this tell us about the broader field? do i need to tie in with the host any more here? (for instance, how microbes might interact with mutated genes/pathways). model, specific microbes on each side, change in mutation or metabolism affects change in community or precedes oncogenesis in a specific way?

@@

#### Sample collection
At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit supplies). At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco Inc.) and ship the samples to the University. Notably, subjects did not undergo any bowel preparation method prior to sampling. On procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System. Patients were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate. A flexible sigmoidoscope was first inserted about 25cm into the colon and endoscopy brush used to collect lumenal/stool contents. Two lumenal samples were collected and the contents immediately deposited into RNAlater (source) and flash-frozen in liquid nitrogen. The brushes were withdrawn and biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter. Three (double check this) mucosal biopsies were collected and flash-frozen in RNAlater. These samples comprised the distal or left colon samples. The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the ascending colon. Samples were then collected as in the distal colon and the colonoscope withdrawn. All samples were stored at -80 C until study completion.

#### Sample processing, sequencing and analysis
DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples. The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (@Kozich 2013). Sequences were curated using the mothur software as described previously (Schloss 2009). The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) (wang 2007) and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff. Sequencing and analysis of a mock community revealed the error rate to be X%. Samples were rarefied to 4231 sequences per sample in order to reduce sampling bias.

Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur (cite). ThetaYC distances were calculated to determine the dissimilarity between two samples. Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location (cite Liaw 2002). To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was >= 99%.

@@

####Figure 6
Location and abundance of cancer-associated OTUs. Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum_ (B) _Porphyromonas asacharolytica_ and (C) _Parvimonas micra_.

###Supplemental material

####Supplemental Figure 1
Random Forest classifier comparing samples to exit stool. ROC curves for the RF model classifying exit stool versus the right or left lumen, pooled mucosa or pooled lumen are shown.

>>>>

---
title: '**Machine learning identifies unique taxa differentiating proximal and distal
  human colonic microbiota**'
fontsize: 11pt
geometry: margin=1.0in
output:
  html_document: default
  pdf_document:
    includes: null
    keep_tex: yes
csl: mbio.csl
bibliography: references.bib 
---
\vspace{35mm}

@@

Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies. A rare report of a matched-colonoscopy study that sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (@Shobar2016). This group found that mucosa and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (@Shobar2016). Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites. Further, the differences were not great enough to overcome interpersonal differences between subjects. So, bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel. Thus our findings of specific bacteria at each site of the colon are strengthened by the lack of bowel preparation.

Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer. Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors. Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit and it performed much better than FIT alone. Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (@Baxter2016). One caveat of the FIT study was that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective. Given that our results showed that the stool most accurately reflects the community of the left lumen, we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.

@@

#### Sample collection
At a baseline visit, subjects were consented and given a home collection stool kit (Source of kit supplies). At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and two swabs of a Fecal Immunohistochemical Test cartridge (Polymedco Inc.) and ship the samples to the University. Notably, subjects did not undergo any bowel preparation method prior to sampling. On procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System. Patients were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate. A flexible sigmoidoscope was first inserted about 25cm into the colon and endoscopy brush used to collect lumenal/stool contents. Two lumenal samples were collected and the contents immediately deposited into RNAlater (source) and flash-frozen in liquid nitrogen. The brushes were withdrawn and biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter. Three mucosal biopsies were collected and flash-frozen in RNAlater. These samples comprised the distal or left colon samples. The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the ascending colon. Samples were then collected as in the distal colon and the colonoscope withdrawn. All samples were stored at -80 C until study completion.

#### Sample processing, sequencing and analysis
DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples. The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described (@Kozich2013). Sequences were curated using the mothur software as described previously (@Schloss2009). The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) (@Wang2007) and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff. Sequencing and analysis of a mock community revealed the error rate to be X%. Samples were rarefied to 4231 sequences per sample in order to reduce sampling bias.

Diversity analysis was performed using the Simpson diversity calculator and theta YC calculator metrics in mothur (@Schloss2009). ThetaYC distances were calculated to determine the dissimilarity between two samples. Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location (@Liaw2002). To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was >= 99%.

@@

####Figure 6
Location and abundance of cancer-associated OTUs. Relative abundance was calculated and plotted by sample site for each OTU of interest: (A) _Fusobacterium nucleatum_ and (B) _Porphyromonas asacharolytica_
